# Continuous Random Variables

**Learning objectives:**

- Learn to define the probability of  continuous event.
- Unify continuous and discrete random variables through
their cumulative distribution functions (**CDF**s)
- What is a Gaussian random variable and why is it 
so important in Data Science?

## Probability Density Function

- The author will denote non countable sets as **continous sets**.

- **What is a PDF?**:
    - It's a continuous version of a PMF.
    - We **integrate** a PDF to compute probability.
    - We integrate instead of [sum](https://math.stackexchange.com/questions/20661/the-sum-of-an-uncountable-number-of-positive-numbers) because *continuous events are not countable*.

  
## More in-depth discussion about PDFs {-}

- **Definition**: A PDF $f_X$ of a random variable $X$ is a mapping
from $\Omega$ to $\mathbb{R}$ with the properties:
    - **Non-negativity**: $f_X \geq 0$ 
    - **Unity**: Integral over sample space equals 1.
    - **Meausre of a set**: $\mathbb{P}[\left\{ x\in A \right\}] = \displaystyle{ \int\limits_{A} f_X\; dx }$ .

- Notice that the PMF mapped $\Omega$ to values in $[0; 1]$, 
but the PDF can map elements in $\Omega$ to any non negative real number.

## Connecting with the PMF {-}

- $f_X(x) = \displaystyle{ \sum_{x_k \in \Omega}} p_X(x_k) \delta(x - x_k)$,
where the **delta** *functions* satisfy:
    - $\delta(x - x_k) = \infty$, if $x = x_k$ 
    - $\delta(x - x_k) = 0, if $x \neq x_k$ 

- Rigorously, we can't integrate such $f_X$ proposed, but, informally,
we can treat a discrete PDF as a train of delta functions.

## Expectation, Moment and Variance

## Definition and properties {-}

- **Definition:** The **expectation** of a continuos random variable $X$ 
is $\mathbb{E}[X] = \displaystyle{ \int_{\Omega} } x \cdot f_X (x)\; dx$.

- The following properties still hold:
    - $\mathbb{E}[cX] = c\cdot \mathbb{E}[X]$
    - $\mathbb{E}[c + X] = c + \mathbb{E}[X]$ 

## Moment and variance {-}

- **Definition:** The **kth moment** of a continuous random variable $X$ is 
$E[X^k] = \displaystyle{ \int_{\Omega} } x^k f_X(x)$.

- **Definition:** The **variance** of a continuous random variable $X$ is
$Var[X] = \mathbb{E}[(X - \mathbb{E}[X])^2] = \displaystyle{ \int\limits_{\omega} (x-\mu)^2 f_X(x) \;dx}$.

- The property $Var[X] = \mathbb{E}[X^2] - \mathbb{E}[X]^2$ still holds.

- **Theorem**: Let $g: \Omega \to \mathbb{R}$ be a function and $X$ be a continuous random variable. Then

$$\mathbb{E}[g(X)] = \int\limits_{\Omega} g(x) f_X(x)\; dx$$

## Existence of expectation {-}

- **Definition**: A random variable $X$ has an expectation if is
**absolutely integrable**:

$$\mathbb{E}[ |X| ] = \int\limits_{\Omega} |x| f_X(x)\; dx < \infty$$

- **Theorem**: Any random variable $X$ satisfies $| \mathbb{E}[X]| \leq \mathbb{E}[|X|]$ .

- **Example**: A random variable whose expectation is undefined is the 
**Cauchy random variable**: $f_X(x) = \dfrac{1}{\pi (1+x^2)}$, for $x\in\mathbb{R}$.

\begin{align}
  \mathbb{E}[|X|] &= \int\limits_{-\infty}^{\infty} |x| \dfrac{1}{\pi (1+x^2)}\; dx \\
  &= 2 \cdot \int\limits_{0}^{\infty} \dfrac{x}{\pi (1+x^2)}\; dx \\
  &\geq 2 \cdot \int\limits_{1}^{\infty} \dfrac{x}{\pi (1+x^2)}\; dx \\
  &\geq 2 \cdot \int\limits_{1}^{\infty} \dfrac{x}{\pi (x^2+x^2)}\; dx \\
  &= \dfrac{1}{\pi}(\log(x))\Bigg{|}_{1}^{\infty}= \infty \;.
\end{align}

## Cumulative Distribution Function

## CDF for continuous random variables {-}

- **Definition**: Let $X$ be a continuous random variable with sample
space $\Omega = \mathbb{R}$. The **cumulative distribution function** (CDF)
of $X$ is

$$F_X(x) = \mathbb{P}[X\leq x] = \int\limits_{-\infty}^{\infty} f_X(x')\; dx'$$

## Properties of CDF {-}

- **Theorem:** Let $X$ be a random variable (continuous or discrete), the its CDF satisfies the following properties:
    1. The CDF is a **non-decreasing**.
    1. The **maximum** of the CDF is when $x=\infty$: $F_X (\infty) = 1$.
    1. The **minimum** of the CDF is when $x=-\infty$: $F_X (\infty) = 0$.

&nbsp; \

- **Definition**: A function $F_X(x)$ is said to be:
    - **Left-continuous** at $x=b$ if $F_X(x) = F_X(b^-) = \lim\limits_{x' \to b^-} F_X(x')$ 
    - **Right-continuous** at $x=b$ if $F_X(x) = F_X(b^+) = \lim\limits_{x' \to b^+} F_X(x')$ 
    - **Continuous** at $X=b$ if it is both left-continuous and right-continuous.

- **Theorem**: The CDF of any random variable (discrete or continuous) is always **right-continuous**.

- **Theorem**: For any random variable $X$ (discrete or continuous):
    -  $\mathbb{P}[X=b] = 0,$ if $F_X$ is continous at $x=b$.
    -  $\mathbb{P}[X=b] = F_X(b) - F_X(b^-),$ if $F_X$ is discontinous at $x=b$.

## Retrieving PDF from CDF {-}

Now is when the fundamental theorem of Calculus (discussed in the first chapter) comes into play.

- **Theorem**: The **probability density function** (PDF) and CDF satisfy:
    - $f_X(x) = \dfrac{d F_X (x)}{dx}$, is $F_X$ is differentiable at $x$ .
    - $f_X(x_0) = \mathbb{P}[X=x_0]\delta(x - x_0)$, is $F_X$ is not differentiable at $x=x_0$ .

## CDF: Unifying discrete and random variables {-}

- The CDF is always a well-defined function, integrable everywhere.
- If the underlying random variable is continuous, so is its CDF.
- If the underlying random variable is discrete, so is its CDF.

## Median, Mode and Mean

We'll know see how to calculate these parameters (commonly used empirically)
for continuous random variables.

## Median {-}

- **Definition**: Let $X$ be a continuous random variable with PDF $f_X$.
The **median** of $X$ is a point $c\in\mathbb{R}$ such that

$$\int\limits_{-\infty}^{\infty} f_X(x) \; dx = \int\limits_{c}^{\infty} f_X(x) \; dx$$

- **Theorem**: The **median** of a random variable $X$ is the point $c$ such that $F_X(c) = 0.5$.

## Mode {-}

- **Definition**: The **mode** of a continuous random variable $X$ is the point
where $f_X$ *attains its maximum*.

## Mean {-}

- The **mean** of a random variable is its **expectancy**.
- **Theorem**: The mean of random variable $X$ can be computed from its
CDF as 

$$
  \mathbb{E}[X] = \int\limits_{0}^{\infty} (1- F_X(t))\; dt 
  - \int\limits_{-\infty}^{0}F_X(t)\; dt
$$

## Uniform and Exponential Random Variables

## Uniform random variables {-}

- **Definition**: Let $X$ be a continuous **uniform random variable** defined in an interval $[a; b]$, then:
    - $f_X(x) = \dfrac{1}{b-a}$, if $x\in [a; b]$.
    - $f_X(x) = 0$, if $x\notin [a; b]$.
    - Notation: $X \sim Uniform(a, b)$ 

- **Theorem**: If $X \sim Uniform(a, b)$, then:
    - $\mathbb{E}[X] = \dfrac{a+b}{2}$.
    - $Var[X] = \dfrac{(b-a)^2}{12}$.

```{r}
# Uniform(0, 1)
x <- seq(-1, 2, 0.01)

plot(
  x, dunif(x, 0, 1), type = 'l',
  main = "PDF", xlab = "x", ylab = ""
)

plot(
  x, punif(x, 0, 1), type = 'l',
  main = "CDF", xlab = "x", ylab = "",
  ylim = c(0,1)
)
```


## Exponential random variables {-}

- **Definition**: Let $X$ be an **exponential random variable** with parameter $\lambda > 0$, then:
    - $f_X(x) = \lambda e^{-\lambda x}$, if $x\geq 0$.
    - $f_X(x) = 0$, if $x<0$.
    - $\lambda$ stands for the **rate of decay** ... larger $\lambda$, faster decay.
    - Notation: $X \sim Exponential(\lambda)$ 

- **Theorem**: If $X \sim Exponential(\lambda)$, then:
    - $\mathbb{E}[X] = \dfrac{1}{\lambda}$.
    - $Var[X] = \dfrac{1}{\lambda^2}$.

- Remember that a Poisson random variable describes the **number of events** 
that happend in a certain **period**. Then, an exponential variable is the **interarrival time** between two consecutive Poisson events; that is, how much time it takes to fo from $N$ Poisson counts to $N+1$ Poisson counts.

```{r}
# Exponential(lambda = 1)
x <- seq(0, 50, 0.01)
lambda <- 1

plot(
  x, dexp(x, 1), type = 'l',
  main = "PDF", xlab = "x", ylab = ""
)

plot(
  x, pexp(x, 1), type = 'l',
  main = "CDF", xlab = "x", ylab = "",
  ylim = c(0,1)
)
```

## Gaussian Random Variables 

- It's also called the **normal random variable**.
- This is the most important continuous random variable, due to its wise use among all scientific disciplines.

- **Definition**: Let $X$ be a **Gaussian random variable** with parameters $\mu, \sigma^2$, then:
    - $f_X(x) = \dfrac{1}{\sqrt{2\pi\sigma^2}}\text{ exp }\left\{ -\dfrac{(x-\mu)^2}{2\sigma^2} \right\}$
    - Notation: $X \sim Gaussian(\mu, \sigma^2) \sim N(\mu, \sigma^2)$ 

- **Theorem**: If $X \sim Gaussian(\mu, \sigma^2)$, then:
    - $\mathbb{E}[X] = \mu$.
    - $Var[X] = \sigma^2$.

## Standard Gaussian {-}

- The PDF of a gaussian random variable does not have an antiderivative
expressable in closed-form (expression that uses a finite amount of standard operations (basic functions and basic arithmetic operations)).

- The common solution is to numerically approximate operations regarding
the CDF of Gaussian(0, 1).

- **Definition**: The **standard Gaussian/normal** random variable 
$X\sim N(0,1)$ has a :
    - PDF: $f_X(x) =\dfrac{1}{\sqrt{2\pi}}e^{-\dfrac{x^2}{2}}$. 
    - CDF: $\Phi(x) = F_X(x) = \dfrac{1}{\sqrt{2\pi}} \displaystyle{ \int\limits_{-\infty}^{x} } e^{-\frac{t^2}{2}}\; dt$ 

- **Theorem**: Let $X \sim \mathcal{N}(\mu, \sigma^2)$, then 
$F_X(x) = \Phi \left( \dfrac{x-\mu}{\sigma} \right)$ 

```{r}
# Gaussian(mean = 0, variance = 1)
x <- seq(-3, 3, 0.01)
mean <- 0
standard_deviation <- 1

plot(
  x, dnorm(x, mean, standard_deviation), type = 'l',
  main = "PDF", xlab = "x", ylab = ""
)

plot(
  x, pnorm(x, mean, standard_deviation), type = 'l',
  main = "CDF", xlab = "x", ylab = "",
  ylim = c(0,1)
)
```


## Skewness and kurtosis {-}

In modern data analysis, some high-order moments usually become useful,
such as **skewness** and **kurtosis**.

- **Skewness**:
    - Measures the **asymmetry** of the distribution.
    - $skewness = \mathbb{E}\left[\left( \dfrac{X-\mu}{\sigma} \right)^3\right] =: \gamma$.
    - Gausian has skewness 0, that is, the distribution is symmetric.

- **Kurtosis**:
    - Measure how **heavy-tailed** the distribution is.
    - $kurtosis = \mathbb{E}\left[\left( \dfrac{X-\mu}{\sigma} \right)^4\right] = \mathcal{k}$ 
    - Gaussian has kurtosis 3.

- **Empirical approximations** 
    - $\gamma \approx \dfrac{1}{N} \displaystyle{ \sum_{n=1}^{N} \left( \dfrac{X_n-\mu}{\sigma} \right)^3}$ 

    - $\mathcal{k} \approx \dfrac{1}{N} \displaystyle{ \sum_{n=1}^{N} \left( \dfrac{X_n-\mu}{\sigma} \right)^4}$ 

## Origin of Gaussian random variables {-}

- Tools for the more formal explanation:

1. The PDF of the **sum of random variables** (X + Y) is the 
**convolution** of $f_X$ and $f_y$ ($f_X * f_y$, which is

$$
  \left( f_X * f_Y \right)(x) = \int\limits_{-\infty}^{\infty} f_X(\tau)
  f_Y(x - \tau) \;d\tau
$$

2. The **Fourier transform** will help us transform convolution into multiplication.

$$
\mathcal{F}\left\{ f_X * \cdots f_X \right\} = 
\mathcal{F}\left\{ f_X \right\} \cdots 
\mathcal{F}\left\{ f_X \right\}\; . 
$$

3. There is a particular sense of convergence, which will be explored later on,
for which the expression $the\;distribution\;of\;the\;sum\;\;X_1 + \cdot X_n\;\;converges\;to\;a\;Gaussian\;distribution$.

## Functions of Random Variables 

- How does the PDF and CDF of a random variable $X$ change
when we apply some function/transformation to $X$?

## General principle {-}

- If the transformation, $g$ is a **one-to-one** mapping, then:
    - $F_Y(y) = \mathbb{P}[Y\leq Y] = \mathbb{P}[g(X)\leq Y] = \mathbb{P}[X\leq g^{-1}(y)] = F_X(g^{-1}(y))$.
    - Via the **chain rule**: $f_Y(y) = \left( \dfrac{d g^{-1}(y)}{dy} \right) f_X(g^{-1}(y))$.

## Generating Random Numbers

- **Random numbers** can be useful to *simulate randomness*, so,
let's see how we can generate them.

## General principle {-}

- We can generate random numbers via an **inverse problem**.

- First, it turns out that is not difficult for a modern
computers to generate random numbers **uniformly, and, between $0$ and $1$** .

- Now, suppose we want to generate random numbers from another random variable $X$, with cummulative distribution function $F_X$.

- Recall that the values of $F_X$ lie between $0$ and $1$.

- Also, not only is $F_X$ a non-decreasing function, for most common cases it's actually **strictly increasing**, which allows us to
correctly define its inverse fuction $F^{-1}_X$.

- Then, we can generate random numbers whose distribution is $f_X$, as follows:
    - Generate a random number $u$ between $0$ and $1$ .
    - Calculate the **unique** value $F^{-1}_X(u)$.
    - **Theorem**: Let $X, Y, U$ be random variables such that
    $U \sim Uniform(0, 1)$ and $Y = F^{-1}_X(U)$. 
    Then, the CDF of $Y$ is $F_X$.

- **Examples**:
    - Generate **Gaussian** random numbers with mean $\mu$ and 
    variance $\sigma^2$.
        - The CDF of the ideal distribution is 
        $F_X(x) = \Phi(\dfrac{x-\mu}{\sigma})$.
        - The transformation becomes
        $F^{-1}_X(U) = \sigma\Phi^{-1}(U) + \mu$.
    - Generate **exponential** random numbers with parameter $\lambda$.
        - The CDF of the ideal distribution is
        $F_X(x) = 1 - e^{-\lambda x}$.
        - The transformation becomes
        $F^{-1}(U) = - \dfrac{1}{\lambda}\log(1-U)$ 

## Meeting Videos {-}

### Cohort 1 {-}

#### Chapter Overview {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>

#### Exercises {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
